---
title: "Maximum Likelihood vs. Bayesian Approaches: Comparative Inference of COVID-19 Vaccine Efficacy"
author: "Casey Lazik, Justin Mai, Harry Schuckman"
date: "`r Sys.Date()`"
output: pdf_document
urlcolor: blue
header-includes:
- \usepackage{amsmath,amsfonts,amssymb}
- \usepackage{setspace} \doublespacing
fontsize: 11pt
---

```{r setup, include=FALSE}
#Use this code chunk to include libraries, and set global options.

library(tidyverse)
library(maxLik)
library(fastR2)
library(LearnBayes)
```

# Abstract

This paper compares the Frequentist and Bayesian approaches to comparative inferences on health data regarding COVID-19 vaccination. The focus centers around the efficacy of the two-dose vaccine (BNT162b2) on SARS-CoV-2 (“COVID-19”), where we use these statistical approaches to conduct hypothesis testing to ensure the vaccine meets the FDA standard of 30 \% efficacy. We are able to conclude this and even apply our models for less forgiving cases like 50 \% efficacy, which is specific to COVID-19 vaccines. Doing so we can see that both the Frequentist and Bayesian approaches yield similar confirming results that lead us to determine that the vaccine demonstrates sufficient efficacy under both criteria.

# Keywords
*Healthcare*, *Efficacy*, *Frequentist*, *Bayesian*

\newpage



# Introduction / Background

This research project was to analyze the efficacy of BNT162b2 COVID-19 Vaccine in comparison to the placebo.

In 2020, when the Coronavirus was at its highest in terms of world impact, scientists were attempting to develop vaccines to reduce the number of COVID-19 cases. To test this, Pfizer researchers randomly appointed candidates of 16 years old or older to either the placebo dosage or the vaccine dosage called BNT162b2, the ratio of placebo vs. vaccine is 1:1. Both parties (doctors and patients) were unaware of which type of they were given/taken to avoid confirmation bias, known as double-blind trials. Each research group were given 2 dosages, 21 days apart, of their respective shots. Then, after at least 7 days since the candidate receives the 2nd dosage, they were tested to see if they had been infected with COVID-19. They stopped conducting this experiment after **170 total infection cases** were seen.

By using the double-blind trials to leave the type of treatment unknown to both the doctor and the patient, it will remove an idea of "suggestion" to avoid subconscious biases. Generally, for drugs such as the BNT162b2 vaccine to be approved by governments and organizations, even though it takes time, it is a necessary research step to improve and test vaccines (Ryding).

The observed number of cases for each treatment group are as follows

\begin{center}
\begin{tabular}{||c c c||} 
 \hline
 Group & Cases & Sample Size \\ [0.5ex] 
 \hline\hline
 BNT162b2 & 8 & 17411 \\ 
 \hline
 Placebo & 162 & 17511 \\
 \hline
 Total & 170 & 34922 \\
 \hline
\end{tabular}
\end{center}

To demonstrate the ratio between the total number of infections for both groups would be $\frac{8}{170} = 0.047$ of cases in the vaccine group out of the total number of infections and $\frac{162}{170} = 0.953$ of cases in the placebo group out of the total number of infections.

In this paper, we intend to use this sample to test the efficacy of this vaccine. More specifically, we will test if this vaccine exceeds the FDA standard of at least 30% efficacy. In order to do this, we will assume the hypothesis that this vaccine does not reach the FDA efficacy standard. We will use this hypothesis because we want to test this vaccine as rigorously is possible due to its importance to the public's health, as evidenced by the FDA's plan to demand 50% efficacy for the COVID-19 vaccine(Sagonowsky). Due to this, we will also include a test for this 50% efficacy figure, although our main focus is on the 30% efficacy standard. As we can see in the figure above, the sample data indicates that the vaccine was effective. The question is as to whether this is enough to disprove our assumed hypothesis. We will be answering this question using both frequentist and Bayesian approaches, and comparing the approaches as well as comparing our results to the Pfizer and BioNTech paper.

```{r label="intro_chart", eval=TRUE, echo=FALSE, fig.height=4, fig.width=4, fig.align='center'}
# You can reference your code in the appendix (sample here).

vac_df <- data.frame(Group = c("BNT162b2", "Placebo", "Total"), 
                     Cases = c(8, 162, 170), 
                     Sample_size = c(17411, 17511, 34922))

vac_df %>% 
  filter(Group == "BNT162b2" | Group == "Placebo") %>% 
  mutate(infected_ratio = Cases/170) %>% 
  ggplot(aes(x = "", y = infected_ratio, fill = Group)) + 
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  ggtitle("BNT162b2 vs Placebo Infection Rate Comparison")
```

# Statistical Methods

## Model



We can model our problem by having T denote the number of people in the COVID-19 group that had the SARS vaccine.


$T\sim Binom(\pi, n=170)$ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ n=170 from data


$\pi = \frac{\pi_v}{\pi_v+ \pi_p}$ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ where $\pi_V$ and $\pi_p$ are the probabilities of an infection in the vaccine and placebo groups respectively.


## Parameters

$\psi=\frac{1-2\pi}{1-\pi}$

$\psi$ represents the efficacy of the two-dose vaccine (BNT162b2). This is how much this vaccine reduces the risk for SARS-Cov-2 (“COVID-19”).




$\pi=\frac{\psi^*-1}{\psi^*-2}$

We then have $\pi$ as the probability of a patient having taken the vaccine given they had COVID-19 (P(Vaccine | COVID-19)).






# *Frequentist Method*




## Confidence Interval

We are assuming that $\widehat{\psi_0^{MLE}}$ follows an approximately normal distribution parameterized by $Norm(\psi_0,\sqrt{ \frac{1}{nI(\psi_0)}  }).$ We make this assumption based on Theorem 12.1(Grove) because our model is a Binomial random variable which is a result of 170 independent Bernoulli trials. Our n=170 is a large enough n that we can assume our MLE $\widehat{\psi_0^{MLE}}$ follows an approximately normal distribution. We use this to set up Wald confidence intervals using this normal distribution.  

$\widehat{\psi_0^{MLE}} \pm z_{\alpha/2} \widehat{SE}$

Where $\alpha = 0.05$, $z_{.025}$ represents the number that cuts off the upper .025 percent of the standard normal PDF, and $\widehat{SE}$ is calculated using the variance from our normal distribution, using our $\psi_0^{MLE}$ value for $\psi_0$.

$\widehat{\psi_0^{MLE}} \pm z_{.025} \sqrt{\frac{1}{nI(\widehat{\psi_0^{MLE}})}}$

See appendix (Proof 1) for calculations.


```{r, echo=FALSE, label="ci_calculation" }
pi_0 = 8/170
psi = (1-2*pi_0)/(1-pi_0)
lower_freq = psi - qnorm(.975)*sqrt(1/(170^2*((pi_0/(1-psi)^2) - (1/(2-psi)^2))) )
upper_freq = psi + qnorm(.975)*sqrt(1/(170^2*((pi_0/(1-psi)^2) - (1/(2-psi)^2))) )
```


We are 95% confident the true value of vaccine efficacy $\psi_0$ lies in the interval [0.9479, 0.9533].

```{r, echo=F, label="taylor_plot", warning=F, fig.height=3, fig.width=5, fig.align='center'}
loglik.binom <- function(psi, n, x){
  if(psi > 1)
    return(NA)
  else{
    loglik = log(choose(n,x)) + x*log(1-psi) - 170*log(2-psi)
  return(loglik)
  }
}

ml.binom <- maxLik2(loglik=loglik.binom, start = 0.3, n=170, x=8)

plot(ml.binom) +
  labs(title = "Second Order Taylor Approximation",
       x=expression(psi))
```


We can be sure that our n is large enough to make the assumptions stated earlier by verifying that the second order Taylor series about the MLE approximates the log-likelihood function.


We also calculated a parametric bootstrap interval for $\psi_0$. This uses our assumption of a binomial distribution with parameter $\pi_0$ and creates new samples from this distribution using simulations, using $n=170$ and our MLE estimation for $\pi_0$ as the parameters. We conducted this parametric bootstrap with 1000 simulations.



## Likelihood Function $L^*(\psi)$


We know $L(\pi)$, since we have a binomial distribution for T. $L(\pi)=\binom{n}{T}\pi^T(1-\pi)^{n-T}$. We used this fact and our parameter for $\pi$ given in terms of $\psi$ to rewrite it as the likelihood function for $\psi$:



$L^*(\psi)=\binom{n}{T}(\frac{\psi-1}{\psi-2})^T(1-\frac{\psi-1}{\psi-2})^{n-T}$ 



## Find MLE of $\psi_0$

Now that we have the likelihood function for $\psi$, we can find the log-likelihood function to be $\ell^*(\psi)=ln(\binom{n}{T}(\frac{\psi-1}{\psi-2})^T(1-\frac{\psi-1}{\psi-2})^{n-T})$. We can then maximize it by taking the derivative and setting it equal to 0. This will produce our MLE equation, which is key for how we go about estimating $\psi$.



$\psi_0^{MLE}=\frac{(n - 2T)}{(n-T)}$  \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ See Proof 2 







## Calculating P-value for likelihood ratio significance test for W


Suppose $H_0: \psi_0 = 0.3$ and $H_1: \psi_0 \ne 0.3$ 



From Chapter 12 slide 22 we have our likelihood ratio test statistic W = 2ln($\Lambda$) where $\Lambda = L(\hat\psi^{mle}_0)/L(\psi^{null}_0)$.

Finding our $\hat\psi^{mle}$ = 0.9506173 we can solve W =  2ln($L(0.9506173)/L(0.3)$) to get our test statistic.

See Proof 3
 

We then use the fact that W $\approx \chi^2_1$   to find P-value = $P(\chi^2_1 \ge 121.6012)$



###  Empirical P-value Calculation

We also calculated the empirical p-value by using our observed W test statistic and simulating 10000 T values from our binomial distribution we have. Using this we get our simulated $\psi$ value and use that alongside our null hypothesis value 0.3 to find the current simulation's likelihood W test statistic. We compute the empirical p-value by the proportion of these W statistics that are at least as large as our observed. 








### Further Calculations

Because we clearly were able to reject for $\psi=0.3$, we wanted to try a harsher test within the context of the FDA requiring even more efficacy specifically for COVID-19 vaccines. Sagonowsky explains that the FDA required at least 50% efficacy for COVID-19 vaccines.



Suppose $H_0: \psi_0 = 0.5$ and $H_1: \psi_0 \ne 0.5$ 

 
P-value = $P(\chi^2_1 \ge 84.4293)$
 









# **Bayesian Methods**

Within the study, it was stated that the prior assumption for parameter $\pi$ is modeled with $Beta(0.700102, 1)$ prior. Also, the vaccine efficacy used is $\psi = \frac{1-2\pi}{1-\pi}$ was estimated to be 95%, this means that by using the Bayesian Inference, they deduced that the vaccine reduces the probability of infection by 95% when compared to the placebo group. Also, the Bayesian credible interval was reported to be [90.3%, 97.6%]. 

*Prior Distribution*: The data model is stated to be a binomial model (vaccine or placebo), which is $T_\pi \sim Binom(170, \pi)$ where $\pi$ is our probability of success P(vaccine|infection).

Our prior distribution is our assumption about the distribution of the probability of an infected person being in the vaccine group. In line with our hypothesis, we will be basing our prior off two assumptions about the vaccine efficacy $\psi$:

$P(\psi \geq 0) = P(\pi \leq \frac{1}{2}) = 0.5$

$P(\psi \geq 0.3) = P(\pi \leq \frac{0.3-1}{0.3-2})=  P(\pi \leq \frac{7}{17})= 0.01$

This ensures that our prior will be centered around an efficacy of 0, and states that in our assumed prior distribution there is only a 1% chance of finding a FDA approved efficacy.

The Beta distribution that satisfies these two inequalities is Beta(`r beta.select(quantile1 = list(p=0.5, x=1/2), quantile2 = list(p=0.01, x=7/17))[1]` , `r beta.select(quantile1 = list(p=0.5, x=1/2), quantile2 = list(p=0.01, x=7/17))[2]`), see prior calculation in appendix.

 
*Posterior Distribution*: The posterior distribution would be given as $Beta(\alpha_{posterior}, \beta_{posterior}) = Beta(\alpha_{prior} + X_V, n - X_V + \beta_{posterior})$ where n is the observed number of cases (170 cases) and $X_V$ is the observed number of cases in the vaccine group (8 cases). 

We are interested in $\psi = \frac{1-2\pi}{1-\pi}$ which is looking at how much the vaccine reduces the risk of infection compared to the placebo. We are also interested in the credible interval for $\psi$ and are testing $H_0: \psi = 0.3, H_a: \psi > 0.3$

*Approach*: To approach this inference, we will then conduct our own Bayesian inference by retrieving the prior and posterior median as our $\pi$ value. We've decided to take this approach because the data sample we are analyzing indicates the probability of receiving an infection with the vaccine is low, meaning we believe the posterior distribution will be strongly right skewed, hence we are using the median instead of the mean because of the impact skewness has on the mean. 

We will then estimate our $\psi$ value based off prior and posterior observations and find the credible interval for $\psi$. Then, we test $H_0: \psi = 0.3, H_a: \psi > 0.3$. We will then analyze our results in comparison to the frequentist approach and the given report.


## Bayesian Inference

For our own inference, we will test the vaccine efficacy and credible interval to compare our results to Pfizer's analysis by utilizing the prior and posterior median using a prior of $Beta(86.04, 86.04)$. To initialize our Bayesian Inference, we will be analyzing the prior assumption, to see the prior probability of receiving an infection in the vaccine group (knowledge of the sample isn't involved).

Figure 3:

```{r label="prior_dist", eval=TRUE, echo=FALSE, fig.height=3, fig.width=3, fig.align='center', message=FALSE}
# Figure 5: prior distribution
pi_values <- seq(0,1,length.out=1000)
prior_values <- dbeta(pi_values, shape1 = 86.04, shape2 = 86.04)

prior_df = data.frame(pi_values, prior_values)

prior_df %>% 
  ggplot(aes(pi_values, prior_values)) + 
    geom_line() +
    geom_vline(xintercept = qbeta(0.5, shape1=86.04, shape2 = 86.04), 
    linetype="dotted") +
    labs(title = "Prior Distribution (Given)")

cat("Probabiliy of receiving an infection in vaccine group:", 
    qbeta(0.5, shape1=86.04, shape2 = 86.04))
```

We will be looking at the median infection probability for the central estimate. This demonstrates that our prior assumption suggests that there is a 0.5 probability of receiving an infection in the vaccine group (based off the median infection probability).

We now observe that in the treatment group, there were 8 infections our of a sample size of 170 (number of total cases). This sets us up for us to find our $Beta(\alpha_{posterior}, \beta_{posterior})$ which is $Beta(\alpha_{prior} + X_V, n - X_V + \beta_{posterior})$ (Grove, Chapter 13.2, Slide 8)





**Hypothesis Testing**: Testing if the vaccine efficacy exceeds 30% based off of the posterior probability using a hypothesis test at $\alpha = 0.05$. We wish to test $H_0: \psi = 0.3, H_a: \psi > 0.3$. We will be testing this using our posterior median

\begin{align*}
  P(\psi \geq 0.3) &= P(\frac{1-2\pi}{1-\pi} \geq 0.3) \\
  &= P(\pi \leq \frac{7}{17}) \\
  P(\pi \leq \frac{7}{17}) &> 0.05 &\text{initialized significance level}
\end{align*}







# Results

We are 95% confident the true value of vaccine efficacy $\psi_0$ lies in the interval [0.9479, 0.9533] using the frequentist method of Wald confidence intervals.


For our data we can compute our Maximum Likelihood estimate (MLE) by using our observed n and t.


$\hat \psi_0^{MLE}=\frac{(170 - 2(8))}{(170-(8))} = 0.9506173$ 


```{r, echo=FALSE, fig.width=6, fig.height=4, fig.align='center'}
set.seed(123)
B = 1000
x <- rbinom(n = B, size = 170, prob = pi_0)
#we are stating that the max likelihood mle of pi = 8/170 (slide 33 chapter 11)
boot_df <- tibble(
  pihat = replicate(n =B,
                    expr = mean(rbinom(n = B, size = 170, prob = pi_0))/170 ),
  psihat = (1-2*pihat)/(1-pihat)
)

ggplot(data = boot_df,
       mapping = aes(x=psihat) )+
  geom_histogram(binwidth = 0.00025)+
  labs(title = "Bootstrapped Sampling Distribution of Psi",
       subtitle = expression(hat(pi)[0] == frac(t,170 )) ,
       x = expression(hat(psi)[0] ==frac(1-2*hat(pi)[0],1-hat(pi)[0] )))

 
```

We have an average of `r round(mean(boot_df$psihat),4)` and standard deviation of 0.000562 for the bootstrap.

The 95 \% confidence interval for the bootstrap is [`r round(quantile(boot_df$psihat, 0.025),3)`,`r round(quantile(boot_df$psihat, 0.975),3)`]


This histogram shows the distribution of $\psi$ over our 1000 bootstrapped samples. We see a bell shaped curve centered around our $\widehat{\psi_0^{MLE}}$ of 0.9506. The bootstrap sampling distribution behaves how we expect the actual sampling distribution to behave, so this suggests that the distribution of $\psi$ is approximately normal and unbiased.  

 



The p-value for the frequentist approach on an efficacy null of 0.3 we found to be $2.822313*10^{-28}$




```{r, label="emp_0.3" ,echo=F, fig.width=6, fig.height=4, fig.align='center'}


B <- 1500
W_sim <- numeric(B)

obs_W <- 121.6012
n <- 170

# Pi value under null of psi=0.3. From pi formula in terms of psi
psi_null <- 0.3
pi_null <- (psi_null - 1 )/(psi_null-2) # pi=(psi-1)/(psi-2) 

loglik <- function(t,psi) {  
  return (log(choose(n, t)) + t * log((psi - 1) / (psi - 2)) +
              (n - t) * log(1 - (psi - 1) / (psi - 2)))
}

set.seed(123)

for (i in 1:B) {
  # Under psi = null
  T_sim <- rbinom(1, 170, pi_null)  # Simulate T from Bin(n=170, pi=pi_null) 
  psi_sim <- (170 - 2*T_sim) / (170 - T_sim)  # Compute MLE for each sample
  
  logL_mle <- loglik(T_sim, psi_sim)
  logL_null <- loglik(T_sim, psi_null)
  
  W_sim[i] <- 2 * (logL_mle - logL_null)
}

empirical_p_value <- mean(W_sim >= obs_W)


```



```{r, echo=F, fig.width=6, fig.height=4, fig.align='center'}

hist(W_sim)

```




Our empirical p-value is `r empirical_p_value` for efficacy 0.3. No simulated W even came close to our observed W.

We additionally found the chi square p-value for testing efficacy of 0.5 to be 3.982026e-20, while also having another 0 empirical p-value.


### Bayesian:

Figure 4:

```{r label="post_dist", eval=TRUE, echo=FALSE, fig.height=3, fig.width=5, fig.align='center', message=FALSE}
# Figure 4: prior and posterior distribution
posterior_values <- dbeta(pi_values, shape1=86.04+8, shape2 = 170-8+86.04)
df = data.frame(pi_values, prior_values, posterior_values)
df %>% 
  ggplot(aes(pi_values)) +
  geom_line(aes(y = prior_values, color = "Prior Distribution")) +
  geom_vline(xintercept = qbeta(0.5, shape1=86.04, shape2 = 86.04), 
             linetype="dotted",
             color = "#00BFC4") +
  geom_line(aes(y = posterior_values, color = "Posterior Distribution")) +
  geom_vline(xintercept = qbeta(0.5, 
                    shape1=86.04+8, 
                    shape2 = 170-8+86.04), linetype="dotted",
                    color = "#F8766D") +
  labs(title = "Prior vs. Posterior",
       x = "Probability of Infection",
       y = "Density",
       color = "Legend")
```



```{r label="post_med", eval=TRUE, echo=FALSE}
# median infection probability for the vaccine group
pi_v_median = qbeta(0.5, 
                    shape1=86.04+8, 
                    shape2 = 170-8+86.04)

# vaccine efficacy
efficacy = (1- 2 * pi_v_median) / (1 - pi_v_median)
```

After observing the data, our posterior probability (updated probability) of receiving an infection in the vaccine group is `r pi_v_median`. Compared to the placebo, the efficacy value based off our posterior median infection probability demonstrates that the vaccine reduces risk of infection calculated using the efficacy equation $\hat{\psi} = \frac{1-2\tilde{\pi}}{1-\tilde{\pi}}$, where $\tilde{\pi}$ represents the posterior median.

```{r label="cred_int", eval=TRUE, echo=FALSE}
pi_v_lower <- qbeta(0.025, shape1=86.04+8, shape2 = 170-8+86.04)
pi_v_upper <- qbeta(0.975, shape1=86.04+8, shape2 = 170-8+86.04)

efficacy_lower <- (1 - 2 * pi_v_upper) / (1 - pi_v_upper)
efficacy_upper <- (1 - 2 * pi_v_lower) / (1 - pi_v_lower)
```

Our estimation for efficacy based off the Bayesian approach using the posterior median of `r pi_v_median` would be `r efficacy`. We found that our 95% posterior interval comes out to `r qbeta(c(0.025,0.975), shape1=86.04+8, shape2 = 170-8+86.04)`. We then apply these values to get the credible interval for efficacy between `r efficacy_lower` and `r efficacy_upper`

After conducting a hypothesis test to test $H_0: \psi = 0.3, H_a: \psi > 0.3$, we found that the posterior probability of $\psi$ being greater than 0.3 is very high (low p-value). Finding p-value using posterior mean we had that given the efficacy is 30%, the likelihood of seeing this value we found the be `r 1-pbeta(7/17, shape1=86.04+8, shape2 = 170-8+86.04)`. See hypothesis testing p-value in appendix.


```{r, echo=F }



# Create results data frame
results <- data.frame(
  Approach = c("Frequentist Approach", "Bayesian Approach"),
  Confidence_Interval_Lower = c(0.9479, round(efficacy_lower,4)),
  Confidence_Interval_Higher = c(0.9533, round(efficacy_upper,4)),
  P_value = c("2.822313e-28", "6.306123e-08")
)

# Print table using kable for basic formatting
knitr::kable(results, caption = "Approach Comparison Table")


```



# Discussion / Conclusion

Based on the results of our analysis, we found that this vaccine meets FDA efficacy standards and should be approved for rollout. According to our confidence intervals, the vaccine should at the minimum have 52% efficacy. Additionally, both the frequentist and Bayesian approaches produces minscule p-values, indicating that it is near impossible for this data to have been produced by a vaccine with less than 0.3 efficacy. Thus, this study has disproved the hypothesis that the vaccine doesn't meet FDA standards.

Using the frequentist approach based on the data from the study, we found that the 95% confidence interval for vaccine efficacy is [0.9479, 0.9533]. Using a bootstrapped sample from this data provided a similar interval of [0.9494, 0.9516]. The Bayesian approach provided a lower efficacy interval of [0.5221, 0.7031]. This is mostly due to the Bayesian approach’s reliance on a prior distribution, the choosing of which can have a significant effect on the results. Pfizer’s analysis chose a relatively weak prior, getting them a credible interval of [90.3%, 97.6%]. We decided to perform our analysis with a stronger prior, strongly stating that the vaccine is not effective in preventing infection. Our prior yields a much lower credible interval in comparison, however this interval is still above 50% efficacy which demonstrates strong evidence of the vaccine's efficacy. Even though the Bayesian approach's credible interval is lower than the frequentist's, all of these confidence intervals support rejecting our hypothesis because all efficacy rates contained in these intervals are beyond the 0.3 standard set by the FDA. Additionally we conducted a p-value calculation against the null hypothesis of an efficacy rate below 0.3, and both frequentist p-values as well as the Bayesian were approximately 0, meaning that it was nearly impossible to achieve the statistics found in the study if the vaccine’s true efficacy was below 0.3 __. It is on the basis of this data that we reject __ the hypothesis that this vaccine’s efficacy is smaller than the FDA’s standard of 0.3. It is clear based on our analysis that this vaccine exceeds FDA efficacy standards.

\newpage

# Bibliography

Fernando P. Polack. (2020). *Safety and Efficacy of the BNT162b2 mRNA Covid-19 Vaccine*. N Engl J Med

Grove, R. (2025). *12: Likelihood Inference*. STAT 342, University of Washington.

Grove, R. (2025). *13.2: Bayes Inference*. STAT 342, University of Washington.

Ryding, Sara. (2021). *What is a Double-Blind Trial?*. Retrieved from https://www.news-medical.net/health/What-is-a-Double-Blind-Trial.aspx

Sagonowsky, Eric. (2020). *FDA will require 50% efficacy for COVID-19 vaccines. How high is that bar?*. Retrieved from https://www.fiercepharma.com/vaccines/fda-to-require-at-least-50-efficacy-for-covid-19-vaccines-wsj

\newpage

# Appendix

Pie Chart: 

```{r ref.label = "intro_chart", eval=FALSE}
```

Proof 1:

$I(\widehat{\psi_0^{MLE}}) = E[-\frac{d^2}{d\psi^2}] \ln(f_{\psi}(t))]$

$\ln(f_{\psi}(t)) = \ln(\binom{170}{t}) + t\ln(1-\psi^*) - 170\ln(2-\psi^*)$

$\frac{d}{d\psi} \ln(f_{\psi}(t)) =  \frac{170}{2-\psi^*} -\frac{t}{1-\psi^*}$

$\frac{d^2}{d\psi^2} \ln(f_{\psi}(t)) = \frac{170}{(2-\psi^*)^2} -\frac{t}{(1-\psi^*)^2}$

$I(\widehat{\psi_0^{MLE}}) = E[-\frac{d^2}{d\psi^2}] = E[\frac{t}{(1-\psi^*)^2} - \frac{170}{(2-\psi^*)^2}] =$

$= \frac{E[t]}{(1-\psi^*)^2} - \frac{170}{(2-\psi^*)^2}$

$= \frac{170\pi_0}{(1-\psi^*)^2} - \frac{170}{(2-\psi^*)^2}$

$SE = \sqrt{\frac{1}{n^2 (\frac{\pi_0}{(1-\psi^*)^2} - \frac{1}{(2-\psi^*)^2})}}$

$\pi_0 = \frac{8}{170}$ using the fact that the MLE of a binomial is $\frac{T}{n}$ (Grove, Chapter 11, Slide 33)

```{r, ref.label="ci_calculation"}


```


Figure 1:

```{r, ref.label="taylor_plot", warning=F, eval=F}



```

 
Proof 2:


$\ell^*(\psi)=ln(\binom{n}{T})+Tln(\frac{\psi-1}{\psi-2})+(n-T)ln(1-\frac{\psi-1}{\psi-2})$


\ 

Minimize $\ell^*(\psi)$ by taking derivative and setting to zero.


$0=\frac{d}{d\psi}(ln(\binom{n}{T})+Tln(\frac{\psi-1}{\psi-2})+(n-T)ln(1-\frac{\psi-1}{\psi-2}))$

$0=\frac{d}{d\psi}(ln(\binom{n}{T})+T(ln(\psi-1)-ln(\psi-2))+(n-T)ln(\frac{\psi-2-\psi+1}{\psi-2}))$

$0=(T/(\psi-1) - T/(\psi-2)+\frac{d}{d\psi}(n-T)ln(\frac{-1}{\psi-2}))$

$0=(T/(\psi-1) - T/(\psi-2)+\frac{d}{d\psi}((n-T)(ln(1)-ln(2-\psi))))$

$0=(T/(\psi-1) - T/(\psi-2)+((n-T)/(2-\psi)))$

$0=(T/(\psi-1) - T/(\psi-2)-(n-T)/(\psi-2))$

$(n-T)/(\psi-2)=T/(\psi-1) - T/(\psi-2)$

$(n-T)=T(\psi-2)/(\psi-1) - T$

$n=T(\psi-2)/(\psi-1)$

$n\psi-n=T\psi-2T$

$n\psi - T\psi=-2T + n$

$\psi_0^{MLE}=\frac{(-2T + n)}{(n-T)}$



Proof 3: 

Calculate L($\hat\psi^{mle}$) = L(0.9506173):

$L(0.9506173) = \binom{n}{T}(\frac{0.9506173-1}{0.9506173-2})^T(1-\frac{0.9506173-1}{0.9506173-2})^{n-T}$

$L(0.9506173) = \binom{170}{8}(\frac{0.9506173-1}{0.9506173-2})^8(1-\frac{0.9506173-1}{0.9506173-2})^{170-8}$

$L(0.9506173) =  0.1429881$


\ 

Calculate L($\psi^{null}$) = L(0.3):


$L(0.3) = \binom{n}{T}(\frac{0.3-1}{0.3-2})^T(1-\frac{0.3-1}{0.3-2})^{n-T}$

$L(0.3) = \binom{170}{8}(\frac{0.3-1}{0.3-2})^8(1-\frac{0.3-1}{0.3-2})^{170-8}$

$L(0.3) =  5.62254*10^{-28}$


\ 

W =  2ln($\frac{0.1429881}{5.62254*10^{-28}}$)

W = 121.6012


We are able to calculate that the observed likelihood ratio test Statistic W for testing $H_0$ (for efficacy 0.3) is 121.6012.

P-value = $P(\chi^2_1 \ge 121.6012)$

```{r}

pchisq(121.6012,1 ,lower.tail=F)

```

\ 

W =  2ln($0.1429881/6.633005*10^{-20}$) 

W = 84.4293


We are able to calculate that the observed likelihood ratio test Statistic W for testing $H_0$ (for efficacy 0.5) is 84.4293.

P-value = $P(\chi^2_1 \ge 84.4293)$

```{r}

pchisq(84.4293,1 ,lower.tail=F)

```


Empirical P-value calculation (efficacy 0.3)


```{r, ref.label="emp_0.3" }




```
```{r}
print(empirical_p_value)
```



Proof 4:



Calculate L($\psi^{null}$) = L(0.5):


$L(0.5) = \binom{n}{T}(\frac{0.5-1}{0.5-2})^T(1-\frac{0.5-1}{0.5-2})^{n-T}$

$L(0.5) = \binom{170}{8}(\frac{0.5-1}{0.5-2})^8(1-\frac{0.5-1}{0.5-2})^{170-8}$

$L(0.5) =  6.633005*10^{-20}$


\ 

Since we calculated L($\hat\psi^{mle}$) = L(0.9506173) = 0.1429881 we have:



W = 2ln($\Lambda$) \ \ \ \ \ \ \ \ \ \ \ \  \ \ \ \ \ \ \ \ \ \ \ \ \  \ \ \ \ \ where $\Lambda = L(\hat\psi^{mle}_0)/L(\psi^{null}_0)$

W =  2ln($0.1429881/6.633005*10^{-20}$) 



Empirical P-value calculation (efficacy 0.5):

```{r}

obs_W <- 84.4293

# Pi value under null of psi=0.5. From pi formula in terms of psi
psi_null <- 0.5
pi_null <- (psi_null - 1 )/(psi_null-2) # pi=(psi-1)/(psi-2) 

```



```{r}

set.seed(123)

for (i in 1:B) {
  # Under psi = null
  T_sim <- rbinom(1, 170, pi_null)  # Simulate T from Bin(n=170, pi=pi_null) 
  psi_sim <- (170 - 2*T_sim) / (170 - T_sim)  # Compute MLE for each sample
  
  logL_mle <- loglik(T_sim, psi_sim)
  logL_null <- loglik(T_sim, psi_null)
  
  W_sim[i] <- 2 * (logL_mle - logL_null)
}

empirical_p_value <- mean(W_sim >= obs_W)
print(empirical_p_value)


```

Figure 3:
```{r ref.label = "prior_dist", eval=FALSE}
```

Figure 4:
```{r ref.label = "post_dist", eval=FALSE}
```

Prior distribution calculation
```{r  eval=FALSE}
beta.select(quantile1 = list(p=0.5, x=1/2),
            quantile2 = list(p=0.01, x=7/17))
```

Posterior median and efficacy calculation
```{r ref.label = "post_med", eval=FALSE}
```

Credible interval calculation
```{r ref.label = "cred_int", eval=FALSE}
```

Hypothesis testing p-value
```{r  eval=FALSE}
# Finding p-value using posterior mean
cat("Given the efficacy is 30%, what is the likelihood of seeing this value:", 
    1-pbeta(7/17, shape1=86.04+8, shape2 = 170-8+86.04))
```
